---
title: "stratR vignette"
author: J. Shekeine
date: "`r Sys.Date()`"
tagline:  <a href="https://github.com/shekeine/stratR"><font size="2">stratR on github</font></a> 
output:
  html_document:
    css: style.css
    toc: yes
    toc_depth: 6
    toc_float:
      collapsed: no
      smooth_scroll: no
---
`r library(knitr)`
`r library(markdown)`
`r library(stratR)`
`r opts_chunk$set(cache=T)`
`r opts_chunk$set(dev.args=list(pointsize=12))`

#Introduction

***  

This vignette illustrates, using practical examples, how to use the `stratR` 
package to construct stratigraphic diagrams of multiple variables. Differences in 
statistical properties amongst variables in large time-series datasets can make it 
difficult visualise such datasets in a way that is accurate, legible and that allows 
simultaneous comparison of all variables of interest. 

`stratR` uses R's `ggplot2` and `gridExtra` libraries to build high quality 
stratigraphic plots of multiple variables that are as well easily customisable. 
It takes care of the numerous fiddly details that would make building such a plot 
a nighmare and provides an intuitive function interface from which key plotting 
parameters can be modified easily. Plotting is done using the `stratPlot` function. 
See `?strapPlot` for full information on the function, its arguments etc.

In addition to the plotting function, `stratR`, through its `hscorer` function, 
provides  a fast way to compute occurence metrics such as percentages and scores 
for paleoenviromental data. Like `stratPlot`, the function was designed with pollen 
stratigraphic data in mind, but can be used on time-series data as well.

***  
##Plotting with stratPlot

###Preparing your data  
***
####Load and describe input data  

Usually, raw data that is to be plotted or analysed in R will be in the [wide](https://en.wikipedia.org/wiki/Wide_and_narrow_data)
format i.e., samples run row-wise and each measured variable is in its own column.
In this vignette, we'll start with a sample dataset that is originally in this format.
Below, we load and inspect a sample dataset from `stratR` that is in the wide format.
```{r}

#Read sample data from stratR
data("stratPlotData")

#Print dimensions
dim(stratPlotData)
```

This dataset contains 125 variables (columns) and 684 samples(rows). We know the 
kind of data is in these columns a priori. For convinience in later steps, we can 
define these variable types as follows:

```{r}
#Metadata columns
metavars <- c('ids', 'entitynum', 'region', 'site',
              'londd', 'latdd', 'altitude', 'samplenum', 'agebp')

#Model output from some model
modvars <- "minD"
  
#Rainfall anomalies
precipvars <- c("JJAPdelta", "DJFPdelta")

#Temperature anomalies
tempvars <- c("JJATdelta", "DJFTdelta")

```

The rest of the columns, 670 in number, contain pollen percentages of 670 plant 
taxa that were observed in various pollen cores, which are uniquley identified 
by the `entitynum` column. Age levels at which pollen counts were observed in each 
pollen core i.e., entitynum, are indicated in the column `agebp`.

It would be too tedious to manually define what columns in the datasets constitute 
pollen counts as done earlier with other variable types. Therefore, we can say 
that every column not already in the metadata, model or climate variable lists 
is a pollen variable.

```{r}
#Pollen variables
pollvars <- names(stratPlotData)[!names(stratPlotData)%in%c(metavars, modvars, precipvars, tempvars)]
```
***
####Reshape to long format  

`stratPlot` expects the input dataset to be in long format. Currently, our input
data is in the wide format. We need to "melt" it to long as follows.
```{r}
datL <- melt(data=stratPlotData, id.vars=metavars, variable.factor=F)
```
***
####Define variable groups
We then need to define the variable types in the data by adding a new column
with grouping labels. Doing this will help determine how different variable
types are aggregated and accorded plotting parameters.

```{r}
#Define variable groups
datL[variable%in%precipvars, group:='Precipitation']
datL[variable%in%tempvars, group:='Temperature']
datL[variable%in%modvars, group:='Model']
datL[variable%in%pollvars, group:='Pollen taxa']
```

The resulting data table now has:

- the original metadata columns

- a column named "variable" with the identities of all measured variables i.e., 
climate, pollen, model variables

- a column named "value" indicating the observed value for the respective measured
variable.

- a column named "group" indicating how the different variables are to be aggregated.

We can get a peek of our reshaped data by running:

```{r, eval=F, echo=T}
head(datL)
```
***
####Remove trace occurences
Since the pollen percentages in the dataset are occurence data, we might want to 
get rid of trace occurences e.g., by retaining only those pollen observations with
rates above a certain thresh-hold, say, 1%. We can also get rid of columns we do 
not wish to plot.

```{r}

#Subset pollen observations with rates equal to or above 1%
poll_dt <- subset(x=datL, variable%in%pollvars & value >=1)

#Fill every sample (time point) where a variable (pollen taxon) was not observed with zero
poll_dt <- dcast(poll_dt,  ids + entitynum + 
                     region + site + londd + latdd + altitude +
                     samplenum + group + agebp ~ variable, value.var='value', fill=0)

#Drop columns with data we dont want to plot, e.g., ones that represent unknown pollen taxa
poll_dt[, NA.:=NULL]

#Reshape pollen data back into the long format
poll_dt <- melt(poll_dt, id.vars=c(metavars, 'group'))

#Remove variables in each entity that do not occur at least once
poll_dt[, n:=sum(value>0), .(entitynum, variable)]
poll_dt <- poll_dt[n>=1]
poll_dt[, n:=NULL]

```

Below, we subset and combine the now ready pollen data subset with the rest of 
the variables.
```{r}

#Subset other non-pollen data i.e. climate and model variables
oth_dt <- subset(x=datL, !variable%in%pollvars)

#Merge processed pollen data with the other non-pollen data
datL <- rbindlist(list(poll_dt, oth_dt), use.names=T)
```
***
###Construct plot
`stratPlot` ingests a dataset of class `data.table` with the aforedescribed 
structure i.e., like `datL` and arguments to control how the output plot is built 
and styled. Details of `statPlot`'s arguments and their meanings can be viewed in 
the package's documentation with `?stratPlot`. Below are some examples illustrating 
the function's arguments and potential use cases.

***
####Single plot
Our sample data currently contains data for 3 entities i.e., pollen cores. The `ids` 
column uniquely identifies each entity i.e., pollen core in the dataset and indicates 
the region associated with each entity. To check what and how many entity `ids` are 
in the dataset we can run:

```{r}
length(unique(datL[, ids]))  #Print number of unique entities
unique(datL[, ids])          #Print labels of unique entities
```

Below, we subset only the data from the entity "Entity_33_Region_Europe", 
which we'll use to generate a sample plot.

```{r}
datL_i <- subset(x=datL, ids=='Entity_33_Region_Europe')
```

The code below generates a stratigraphic plot of this subset. See `?stratPlot`
for details on the arguments.

```{r, eval=T}

#Build plot
plt_ls <- stratPlot(dat=datL,
                   grpcol='group',
                   tymcol='agebp', varcol='variable', valcol='value', varord=c('med'),
                   grp_ord=c('Model', 'Precipitation', 'Temperature', 'Pollen taxa'),
                   grp_colours=c('grey40', 'blue', 'brown3','darkgreen'),
                   grp_geom=c('line', 'line', 'line','area'),
                   grp_bline=c(NA, 20, "med", "med"),
                   grp_vartype=c('MinD', 'SD', 'SD','Percentages'),
                   grp_const=c('Model', 'Precipitation', 'Temperature'), 

                   xbrkint=c(NA, NA, 5, NA),
                   ybrkint=NA,

                   axs_tsize=10, var_tsize=18, grp_tsize=25, ptt_tsize=20,

                   plab='', ylab='Age BP',
                   pmargin=unit(c(0.05, 0, 0.1, 0), 'cm'),
                   maxw=5
                  )

#Get plot object
plot_obj <- plt_ls$plot

#Get plot width (in cm)
plot_width <- plt_ls$width

```

The constructed plot can be viewed on the R graphics device in two quick lines.

########{#wrapper}
```{r, eval=T, fig.height=25/2.54, fig.width=plot_width/2.54}

#Draw plot
grid.newpage()
grid.draw(plot_obj)

```

#######{#endCss}

To write the plot to file for later viewing or use in articles etc, we can use 
the `ggsave` function from `ggplot2` . To save to other formats, just change the 
extension of the output file name to the desired type e.g., png, tif, etc. Here, 
the plot is saved to the .pdf format.

```{r, eval=T}

#Save plot as pdf
ggsave(filename=paste0("/tmp/", 'myplot.pdf'), plot=plot_obj,
       limitsize=F, width=plot_width, height=25, units='cm')

```

***
####Multiple plots
To construct plots for multiple entities i.e., pollen cores, we can call `stratPlot` 
from within a `for loop` that will build a collection stratigraphic plots: one for 
each entity. These could be useful in various ways e.g., in comparing different 
entities, or, in generating a library of plots that can be used in an interactive 
map to query and view data at different locations.

***
#####Using sequential for loops
Our main dataset contains 3 entities.

```{r, eval=T}
length(unique(datL[, ids]))  #Print number of unique entities
unique(datL[, ids])          #Print labels of unique entities
```

Each entitity's data can be visualised using stratPlot with an ordinary for loop 
as follows:

```{r, eval=T}
  for (entity in unique(stratPlotData[, ids])){
    
    #Isolate data for entity in the current iteration
    dat_ent <- datL[ids==entity]
    
    #Build plot
    plt_ls <- stratPlot(dat=dat_ent,
                       grpcol='group',
                       tymcol='agebp', varcol='variable', valcol='value', varord=c('med'),
                       grp_ord=c('Model', 'Precipitation', 'Temperature', 'Pollen taxa'),
                       grp_colours=c('grey40', 'blue', 'brown3','darkgreen'),
                       grp_geom=c('line', 'line', 'line','area'),
                       grp_bline=c(NA, 20, "med", "med"),
                       grp_vartype=c('MinD', 'SD', 'SD','Percentages'),
                       grp_const=c('Model', 'Precipitation', 'Temperature'), 
    
                       xbrkint=c(NA, NA, 5, NA),
                       ybrkint=NA,
    
                       axs_tsize=10, var_tsize=18, grp_tsize=25, ptt_tsize=20,
    
                       plab=entity, ylab='Age BP',               #Note plab value here
                       pmargin=unit(c(0.05, 0, 0.1, 0), 'cm'),
                       maxw=5
                      )
    
    #Get plot object
    plot_obj <- plt_ls$plot
      
    #Get plot width (in cm)
    plot_width <- plt_ls$width
    
    #Make filename, replace "/tmp/" with the path you want your plots saved
    fname <- paste0("/tmp/", entity, ".pdf")
    
    #Save plot as pdf
    ggsave(filename=fname, plot=plot_obj,
           limitsize=F, width=plot_width, height=25, units='cm')

}

```

***
#####Using parallel for loops
In addition to the ordinary for loops, we can also exploit R's parallel processing 
capabilities to speed up the generation of multiple plots. Note that 
to construct just one or two plots, an ordinary (sequential) `for loop`  should be 
preffered for reasons explained [here](http://stackoverflow.com/questions/14614306/why-is-the-parallel-package-slower-than-just-using-apply.)
As the number of plots to construct increases, say, beyond 5, running the task in 
parallel becomes a lot more efficient compared running the task sequanetially as 
the standard `for` loop does in R.

```{r, eval=T}
  #Load libraries for parallel processing
  library('parallel')
  library('foreach')
  library('doParallel')

  #Register cluster for parallel processing
  cl <- makeCluster(detectCores() - 1)
  registerDoParallel(cl, cores = detectCores() - 1)
  
  #Parallel for loop
  plotList <- foreach (entity=unique(stratPlotData[, ids]), .packages="stratR", .combine='c') %dopar% {
    
    #Isolate data for entity in the current iteration
    dat_ent <- datL[ids==entity]
    
    #Build plot
    plt_ls <- stratPlot(dat=dat_ent,
                       grpcol='group',
                       tymcol='agebp', varcol='variable', valcol='value', varord=c('med'),
                       grp_ord=c('Model', 'Precipitation', 'Temperature', 'Pollen taxa'),
                       grp_colours=c('grey40', 'blue', 'brown3','darkgreen'),
                       grp_geom=c('line', 'line', 'line','area'),
                       grp_bline=c(NA, 20, "med", "med"),
                       grp_vartype=c('MinD', 'SD', 'SD','Percentages'),
                       grp_const=c('Model', 'Precipitation', 'Temperature'), 
    
                       xbrkint=c(NA, NA, 5, NA),
                       ybrkint=NA,
    
                       axs_tsize=10, var_tsize=18, grp_tsize=25, ptt_tsize=20,
    
                       plab=entity, ylab='Age BP',               #Note plab value here
                       pmargin=unit(c(0.05, 0, 0.1, 0), 'cm'),
                       maxw=5
                      )
    #Get result
    res <- list(plt_ls) 
    names(res) <- entity
    return(res)
  }
  stopCluster(cl)
```

The object `plotList` created above is a nested list object containing a plot object 
for each entity processed as well as the respective width. Below, we access each 
plot object in the list and save it to disk at the respective width.

```{r, eval=T}

 #Write plot objects to file  
 for (entity in names(plotList)){
  
  #Get plot object in current iteration
  plot_obj <- plotList[[entity]]$plot
    
  #Get plot width (in cm)
  plot_width <- plotList[[entity]]$width
  
  #Make filename, replace "/tmp/" with the path you want your plots saved
  fname <- paste0("/tmp/", entity, "_par_.pdf")
  
  #Save plot as pdf
  ggsave(filename=fname, plot=plot_obj,
         limitsize=F, width=plot_width, height=25, units='cm')
 }
```

***
##Computing biome, pft scores with hscorer
***
The function `hscorer` computes pollen scores based on biome, PFT or other schemes.
Though built for pollen data, it can be used or adapted for use to calculate 
occurence metrics for other paleoenvironmental data as well.

####Preparing data
This function's documentation (see `?hscorer`) explains the expected inputs datasets 
and what structure they should have. Here, we'll get straight to work. `stratR` 
contains sample datasets that can be used as input for `hscorer` We can load them 
as follows:

```{r}
#Load sample pollen counts
data(p_counts) #Load a sample pcounts table (data.table of pollen counts)
head(p_counts) #Inspect: This dataset correponds to the expected value of the `pcounts` argument
```

Below, we load sample biomisation and pft schemes. Note that the object `hschemes` 
being loaded below is a collection of several schemes (several tables)
```{r, echo=F, eval=F}
#Load sample pft and biomisation schemes
data(hschemes)         #Load a bunch of sample pft and biomisation schemes
print(hschemes)        #Each table correponds to the expected value of the `hscheme` argument
print(names(hschemes)) #See what biomisation schemes are available
```

***
####Single scheme
Calculating biome, pft or any other kind of occurence metrics using a single scheme 
is as simple as laid out below. Note that here, we have a collection of schemes so 
the first line just gets us the one we want from the whole bunch e.g. the 
Peyron biome scheme.

```{r, echo=F, eval=F}
    #Isolate peyron biome scheme
    peyron_bs    <- hschemes$Peyron_biome_scheme

    #Calculte biome scores, and by default, percentages
    percs_scores <- hscorer(pcounts=p_counts, hscheme=peyron_bs)
    
    #Inspect output
    print(percs_scores)
```

As outlined in `hscorer` documentation (see`?hscorer`) the output value, 
`percs_scores` in this case, contains percentages and scores for the pollen data 
as per the scheme used: Peyron's biome scheme in this example.

***
####Multiple schemes
Occurence metrics can also be computed based on multiple schemes by using a `for` 
loop that iterates over the different schemes we have, outputting a result based 
on each scheme. The snippet below illustrates a use case where we want to generate 
occurence metrics based on several schemes.

```{r}

for(scheme in names(hschemes)){
    
    #Get i'th bscheme
    hscheme_i <- hschemes[[scheme]]
    
    #Compute occurence percentages and scores based on i'th scheme
    percs_scores_i  <- hscorer(pcounts=p_counts, hscheme=hscheme_i)
    
    #Make filenames, replace "/tmp/" with the path where you want your plots saved
    fname_percs <- paste0("/tmp/", "percentages_", scheme, ".csv")
    fname_scores <- paste0("/tmp/", "scores_", scheme, ".csv")
    
    #Write percentages to file
    write.csv(x=percs_scores_i$percentages, file=fname_percs, row.names=F)
    
    #Write occurence scores to file
    write.csv(x=percs_scores_i$scores, file=fname_scores, row.names=F)
  }
```

***
####Plotting output with stratPlot
In conclusion, outputs from `hscorer` can be readily visualised with the `stratPlot` function. 
Below, we compute PFT percentages based on Peyron`s PFT scheme and plot the result 
with `stratPlot`.

```{r}

#Isolate peyron pft scheme
peyron_pft <- hschemes$Peyron_pft_scheme

#Compute taxon percentages
peyron_percs <- hscorer(pcounts=p_counts, hscheme=peyron_pft)$percentages

#Get the data ready for stratPlot

  #Subset data for just one entity
  dat <- peyron_percs[entitynum==221]
  
  #Melt to long format
  datm <- melt(dat, id.vars=c("entitynum", "samplenum"), variable.name="PFT")
  
  #Remove trace occurences (less than 1% rate)
  datm <- datm[value>1]
  
  #Fill every sample (time point) where a PFT was not observed with zero
  datm <- dcast(datm,  entitynum + samplenum ~ PFT, value.var='value', fill=0)
  
  #Reshape pollen data back into the long format
  datm <- melt(datm, id.vars=c("entitynum", 'samplenum'), variable.name="PFT")
```

We now need to add a grouping variable as expected by `stratPlot`. For demo 
purposes, we'll just lump all the PFT's in one artificial group. However, one 
could, for example, add biome data from a pft to biome translation table as the 
grouping variable so that the time-series of PFT percentages are plotted by biome.

```{r}
datm[, group:="Plant functional types"]
```

Finally, we can construct the plot as illustrated earlier as follows.

```{r, eval=T}
#Plot
plt_ls <- stratPlot(dat=datm,
                       grpcol='group',
                       tymcol='samplenum', varcol='PFT', valcol='value', varord=c('med'),
                       grp_geom=c('area'),
                       grp_bline=0,
                       grp_vartype="Percentages",
                       grp_colours="darkgreen",

                       axs_tsize=10, var_tsize=18, grp_tsize=16, ptt_tsize=18,
    
                       plab="Entity 506", ylab='Sample',
                       pmargin=unit(c(0.05, 0, 0.1, 0), 'cm'),
                       maxw=5
                      )
#Get plot object
plot_obj <- plt_ls$plot

#Get plot width (in cm)
plot_width <- plt_ls$width
```

The generated plot can then be on the R device as follows.

########{#wrapper2}
```{r, eval=T, fig.height=20/2.54, fig.width=plot_width/2.54}

#Draw plot
grid.newpage()
grid.draw(plot_obj)

```

#######{#endCss}
